---
title: "More Examples of Time-Dependent Cox Model"
author: "Yang Qiao"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(survival)
library(survminer)
options(width = 100)
```

# More Examples of Time-Dependent Cox Model

This page provides two more complicated examples of time-dependent Cox models.
In both examples, we use simulated data where certain covariates may change during follow-up.

- The 1st example looks at one time-varying covariate (Tvc) together with age and gender;
- The 2nd example includes two time-varying covariates at the same time.

For each case, we fit a Cox model and show predicted survival curves, comparing situations where the Tvc changed versus remained unchanged.

## Data Generation

We simulate a small cohort with two fixed covariates (`age` and `gender`) and two time-varying covariates:

- Tvc1 (`X1`) tends to **increase** the hazard (risk of the event) when it happens.
- Tvc2 (`X2`) tends to **decrease** the hazard when it happens.
- The earlier these occur, the **stronger** their effect on hazard.

What this means

- `event = 1` means we observed the event (e.g., death) by time.
- `X1t_time` and `X2t_time` are the days when that time-varying covariate changed. If NA, it didn't change before the observed time.
- We made Tvc1 (`X1`) harmful on average (increases risk) and Tvc2 (`X2`) protective on average (reduces risk), with earlier times having stronger impact.

```{r}
set.seed(1)
n <- 50
time_censor <- 160

data <- tibble(id = paste0("s", 1:n)) %>%
  mutate(
    # (1) Add fixed (non time-varying) covariates
    age = sample(30:80, n(), replace = T),
    gender = sample(c("M", "F"), n(), replace = T),
    
    # (2) Add time-varying covariates X1t and X2t
    X1t_value = rbinom(n(), 1, 0.6),
    X1t_time  = if_else(X1t_value == 1, round(runif(n(), time_censor * 0.1, time_censor * 0.6)), NA),
    X2t_value = rbinom(n(), 1, 0.6),
    X2t_time  = if_else(X2t_value == 1, round(runif(n(), time_censor * 0.1, time_censor * 0.6)), NA),
    
    # (3) Add event-time response from Weibull distribution
    base_rate   = 0.005 * log10(age),
    effect_X1   = exp(1 * X1t_value * replace_na(1 - X1t_time / time_censor, 0)),
    effect_X2   = exp(-1 * X2t_value * replace_na(1 - X2t_time / time_censor, 0)),
    adj_rate    = base_rate * effect_X1 * effect_X2,
    true_time   = ceiling(rweibull(n(), shape = 2, scale = 1 / adj_rate)),
    true_censor = sample((time_censor / 2):time_censor, n(), replace = T),
    time        = pmin(true_time, true_censor),
    event       = as.integer(true_time <= true_censor),
    
    # (4) Adjust X1t and X2t if after observed time
    X1t_time   = if_else(!is.na(X1t_time) & X1t_time >= time, NA, X1t_time),
    X1t_value  = if_else(is.na(X1t_time), 0, X1t_value),
    X2t_time   = if_else(!is.na(X2t_time) & X2t_time >= time, NA, X2t_time),
    X2t_value  = if_else(is.na(X2t_time), 0, X2t_value)
  ) %>%
  select(id, event, time, age, gender, X1t_value, X1t_time, X2t_value, X2t_time)

# Quick data check
summary(data)
print(data, n = Inf)
```


## 1. Cox model with one time-varying covariate and multiple fixed covariates
```{r}
# Build start–stop data with a single time-varying covariate tvc (for X1)
data_td <- data %>%
  tmerge(., ., id = id, event = event(time, event), tvc = tdc(X1t_time, X1t_value)) %>%
  mutate(across(tvc, ~ replace_na(., 0)))
# Fit Cox model: hazard ~ X1(t) + age + gender
model <- coxph(Surv(tstart, tstop, event) ~ tvc + age + gender, data = data_td, id = id, timefix = F)
summary(model)
```

How to read the model

- The coef for `tvc` is the log-hazard ratio when Tvc1 (`X1`) changed.
- exp(coef) = hazard ratio (HR):
- HR > 1: higher instantaneous risk (shorter survival) when Tvc1 (`X1`) changed.
- HR < 1: lower instantaneous risk (longer survival) when Tvc1 (`X1`) changed.
- p-value: evidence that Tvc1 (`X1`) affects hazard, after adjusting for age/gender.

Below we create two scenarios for prediction: Tvc1 unchanged vs Tvc1 changed at a typical time (its median time), holding age at its median and comparing by gender.
Note these are model-based survival curves (not raw Kaplan–Meier).

```{r}
# Create a new data to visualize predicted survival
data_sim <- data.frame(event = 1:0, Xt_time = c(NA, median(data_td$X1t_time, na.rm = T)), Xt_value = 0:1) %>%
  # Use the maximum follow-up time for prediction horizon
  mutate(time = max(data_td$time), .before = 1) %>%
  # Compare by gender at median age
  expand_grid(gender = c("M", "F"), age = median(data_td$age)) %>%
  mutate(id = paste0("s", 1:n()), .before = 1)
data_sim

# Convert to start–stop format
data_fit <- data_sim %>%
  tmerge(., ., id = id, event = event(time, event), tvc = tdc(Xt_time, Xt_value)) %>%
  mutate(across(tvc, ~ replace_na(., 0)))
data_fit

# Fit model-based survival for each individual
fit <- survfit(model, newdata = data_fit, id = id)

# Plot: model-based predicted survival curves
label <- with(data_sim,
              sprintf("Gender=%s, Age=%d, Tvc1 %s", gender, age, if_else(Xt_value == 1, paste("changed at time", Xt_time), "unchanged")))
pp <- ggsurvplot(fit, data_fit, censor = F, legend.labs = label, font.legend = 9)
pp$plot + labs(color = NULL) + guides(color = guide_legend(nrow = 2))
```

## 2. Cox model with two time-varying covariates
```{r}
# Build start–stop data with two time-varying covariates tvc1 (for X1) and tvc2 (for X2)
data_td <- data %>%
  tmerge(., ., id = id, event = event(time, event), tvc1 = tdc(X1t_time, X1t_value), tvc2 = tdc(X2t_time, X2t_value)) %>%
  mutate(across(c(tvc1, tvc2), ~ replace_na(., 0)))
# Fit Cox model: hazard ~ X1(t) + X2(t)
model <- coxph(Surv(tstart, tstop, event) ~ tvc1 + tvc2, data = data_td, id = id, timefix = F)
summary(model)
```

How to read the two-tvc model

- `tvc1` HR > 1 means Tvc1 (`X1`) changed increases hazard; HR < 1 means it protects.
- `tvc2` HR > 1 means Tvc2 (`X2`) changed increases hazard; HR < 1 means it protects.
- Signs and magnitudes are adjusted for each other.

We compare the four combinations:

- Tvc1 and Tvc2 unchanged.
- Tvc1 unchanged and Tvc2 changed at its median time.
- Tvc1 changed at its median time and Tvc2 unchanged.
- Tvc1 and Tvc2 changed at their own median time.

```{r}
# Create a new data to visualize predicted survival
data_sim <- expand_grid(
  data.frame(X1t_time = c(NA, median(data_td$X1t_time, na.rm = T)), X1t_value = 0:1),
  data.frame(X2t_time = c(NA, median(data_td$X2t_time, na.rm = T)), X2t_value = 0:1)
) %>%
  # Use the maximum follow-up time for prediction horizon
  mutate(id = paste0("s", 1:n()), time = max(data_td$time),
         event = c(1L, rep(0L, n() - 1)),
         .before = 1)
data_sim

# Convert to start–stop format
data_fit <- data_sim %>%
  tmerge(., ., id = id, event = event(time, event), tvc1 = tdc(X1t_time, X1t_value), tvc2 = tdc(X2t_time, X2t_value)) %>%
  mutate(across(c(tvc1, tvc2), ~ replace_na(., 0)))
data_fit

# Fit model-based survival for each individual
fit <- survfit(model, newdata = data_fit, id = id)

# Plot: model-based predicted survival curves
label <- with(data_sim,
              sprintf("Tvc1 %s, Tvc2 %s", if_else(X1t_value == 1, paste("changed at time", X1t_time), "unchanged"),
                      if_else(X2t_value == 1, paste("changed at time", X2t_time), "unchanged")))
pp <- ggsurvplot(fit, data_fit, censor = F, legend.labs = label, font.legend = 9)
pp$plot + labs(color = NULL) + guides(color = guide_legend(nrow = 2))
```
